A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-02-18 analysis/paper/paper.Rmd
update figure labels and result section after rerunning models

<!--archaeological specific case study and our model-->
We use a Bayesian ERGM to study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan (Figure \@ref(fig:figure-KWL-map)), which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. Our general model proposes that the influence of a colonial power, combined with high local values attached to imported goods, leads to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. Increased social inequality associated with the use of foreign prestige goods is recognized as network strategies for gaining power [@Blanton1996; @Feinman2000]. Network-based societies are identified in some places in Asia with involvment in long-distance trade [@Carter2015; @Ueda2016; @Liu2006]. The observed uneven distribution of burial goods at Kiwulan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by Europeans.

@@

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For nodal covariates in the pre-European model, the ritual element represent by pots (ritual-homophily) has a significant effect on the formation of relations between burials. On the other hand, the wealth rank, represented by burial values (wealth-homophily), has a negative effect. This demonstrates that burials in the pre-European period with ritual pottery tend to form relations, but burials in the same wealth level tend not to. Despite positive mean values for some covariates, such as age and sex, their confidence intervals include zero, so they do not show a significant influence. Similarly, the dyadic covariate, physical distance, shows no significant effect, indicating that physical proximity between burials does not reflect similarity in burial treatment. For the endogenous network effects, transitivity presents a significant positive effect, while centralization demonstrates a negative effect. The high positive value for transitivity suggests a tendency of burials with similar burial goods to be clustered as connected communities, indicative of the presence of multiple corporate groups sharing burial goods in common. In contrast, the strong negative centralization shows there is a tendency toward decentralization that reflects most burials having a similar number of ties, without any prominent burials. This might imply that individuals have equal access to the flow of trade goods.

```{r table-posterior-stats}
# tidy data for the table of posterior stats 
posterior_esti_two_phases<-
  bergm_two_phases %>%
  pivot_longer(!phase, 
               names_to = "parameter", 
               values_to = "posterior") %>%
  mutate(parameter= case_when(parameter == "edges" ~ "Density",
                              parameter == "nodematch.age" ~ "Age-homophily",
                              parameter == "nodematch.sex" ~ "Sex-homophily",
                              parameter == "nodematch.ritual" ~ "Ritual-homophily",
                              parameter == "nodematch.value" ~ "Wealth-homophily",
                              parameter == "gwesp" ~ "transitivity",
                              parameter == "gwdeg" ~ "centralization",
                              parameter == "dyadcov.distance" ~ "Physical distance")) %>% 
  mutate(parameter = factor(parameter,
                            levels=c("Density", "Age-homophily", 
                                     "Sex-homophily", "Ritual-homophily",
                                     "Wealth-homophily", "transitivity",	
                                     "centralization", "Physical distance"))) %>% 
  group_by(parameter, phase) %>% 
  summarize(mean = round(mean(posterior), 2),
            median = round(quantile(posterior, 0.5), 2),
            `2.5%` = round(quantile(posterior, 0.025), 2),
            `97.5%` = round(quantile(posterior, 0.975), 2)) %>% 
  pivot_wider(names_from = phase, 
              values_from = c(mean, median, `2.5%`, `97.5%`)) %>% 
  select(parameter, ends_with(c("pre", "post")))

@@

```{r figure-posterior-distribution, fig.cap = "Posterior density estimates for the parameters associated with edges, gwesp, gwdegree, and distance-dyadcov by phases. The pre-European group presents remarkable larger values for the gwesp, but lower values for the edges and gwdegree parameters compared to the post-European group. The distance parameters overlap between two groups. Both posterior estimates present generally symmetric distributions."}
knitr::include_graphics(here::here("analysis", "figures", "006-posterior-distribution.png"))
```

<!--posterior stats for for post-E and their comparison-->
After the arrival of Europeans, the nodal covariates of wealth rank show significant positive effects, while the ritual and sex variables have negative effects. This indicates that burials in the same wealth level tend to be more connected with each other. On the contrary, the same sex burials, and burials with ritual pottery in common tend, not to have relations. In the post-European burials there are no significant effects for variables of age and physical distance. Similar to the pre-European network, transitivity demonstrates a significant positive effect, but much weaker than in the pre-European network. In contrast, centralization has a significantly higher positive effect than the pre-European network. This means there is a tendency toward centralization that reflects a limited number of burials having many more ties than others. This implies more access to trade goods and the increased wealth accumulation and display behaviors in burial events. In general, the post-European network model has a smaller transitivity effect and a positive centralization effect compared to the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence.

@@

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand the robustness of comparison between two networks, we used a vertex bootstrap technique to cross-validate the results of our Bayesian ERGMs. The vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) to quantify standard errors and estimate sampling variability in the network statistics of interest [@Chen2019; @Snijders1999; @Rroberts2021]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks of different sizes by examining their confidence intervals for the network population. We used the vertex bootstrap to compute endogenous network statistics, including density, centralization, and transitivity for our two networks. We explored the sample size effect by removing nodes at certain percentages (5-40%) for both networks and comparing their confidence intervals. The results demonstrate a consistent difference between the two networks up to 30% node removals (Figure \@ref(fig:figure-bootstrap-CI)). This indicates that the network variables under investigation are robust to the different network size of our sample. Figure \@ref(fig:figure-bootstrap-CI also shows consistency with our earlier finding of negative centralization in the pre-European period and positive centralization in the post-European period, over several conditions of node removal. Similarly consistent with our earlier result, the vertex bootstrap shows no significant difference on density and transitivity.

@@

<!--archaeological implication based on the results-->
Network covariates give insights into the socio-cultural factors that were associated with the acquisition and distribution of foreign prestige goods. Our results indicate that the major factor contributing to burial networks is ritual for the pre-European network, and wealth for the post-European network. We interpret these results as illustrating different social and economic mechanisms influencing relations between burials for the two periods. Before European arrival, ritual behaviors were important factors that structured the interconnections of prestige goods in the burial contexts. This might imply that ritual could be a status indicator at the time before foreign goods were first introduced to Indigenous societies. After the European presence was established, status indicators shifted from ritual to wealth differences. Moreover, geographical proximity, implying some kinship association, has not effect on sharing similar prestige items between burials for both phases. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based. Demographic information, such as age and sex, do not have any effects on network formation for the pre-European network and only sex shows a negative effect for the post-European network. This may also support a shift to more wealth and individual-based status competition without sex differences, in the later phase. But we also note that no effects could be due to the unavailability of age and sex values for many burials in our sample.

>>>>

<!--archaeological specific case study and our model-->
We use a Bayesian ERGM to study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan (Figure \@ref(fig:figure-KWL-map)), which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. Our general model proposes that the influence of a colonial power, combined with high local values attached to imported goods, leads to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. Increased social inequality associated with the use of foreign prestige goods is recognized as network strategies for gaining power [@Blanton1996; @Feinman2000]. Network-based societies are identified in some places in Asia with involvement in long-distance trade [@Carter2015; @Ueda2016; @Liu2006]. The observed uneven distribution of burial goods at Kiwulan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by Europeans.

@@

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For nodal covariates in the pre-European model, the ritual element represent by pots (ritual-homophily) has a significant effect on the formation of relations between burials. On the other hand, the wealth rank, represented by burial values (wealth-homophily), has a negative effect but not significant since its confidence intervals include zero. This demonstrates that burials in the pre-European period with ritual pottery tend to form relations. Despite positive mean values for some covariates, such as age and sex, they do not show a significant influence. Similarly, the dyadic covariate, physical distance, shows no significant effect, indicating that physical proximity between burials does not reflect similarity in burial treatment. For the endogenous network effects, transitivity presents a significant positive effect, while centralization demonstrates a negative effect. The high positive value for transitivity suggests a tendency of burials with similar burial goods to be clustered as connected communities, indicative of the presence of multiple corporate groups sharing burial goods in common. In contrast, the strong negative centralization shows there is a tendency toward decentralization that reflects most burials having a similar number of ties, without any prominent burials. This might imply that individuals have equal access to the flow of trade goods.

```{r table-posterior-stats}
# tidy data for the table of posterior stats 
posterior_esti_two_phases<-
  bergm_two_phases %>%
  pivot_longer(!phase, 
               names_to = "parameter", 
               values_to = "posterior") %>%
  mutate(parameter= case_when(parameter == "edges" ~ "Density",
                              parameter == "nodematch.age" ~ "Age-homophily",
                              parameter == "nodematch.sex" ~ "Sex-homophily",
                              parameter == "nodematch.ritual" ~ "Ritual-homophily",
                              parameter == "nodematch.value" ~ "Wealth-homophily",
                              parameter == "gwesp" ~ "Transitivity",
                              parameter == "gwdeg" ~ "Centralization",
                              parameter == "dyadcov.distance" ~ "Physical distance")) %>% 
  mutate(parameter = factor(parameter,
                            levels=c("Density", "Age-homophily", 
                                     "Sex-homophily", "Ritual-homophily",
                                     "Wealth-homophily", "Transitivity",	
                                     "Centralization", "Physical distance"))) %>% 
  group_by(parameter, phase) %>% 
  summarize(mean = round(mean(posterior), 2),
            median = round(quantile(posterior, 0.5), 2),
            `2.5%` = round(quantile(posterior, 0.025), 2),
            `97.5%` = round(quantile(posterior, 0.975), 2)) %>% 
  pivot_wider(names_from = phase, 
              values_from = c(mean, median, `2.5%`, `97.5%`)) %>% 
  select(parameter, ends_with(c("pre", "post")))

@@

```{r figure-posterior-distribution, fig.cap = "Posterior density estimates for the parameters associated with edges, transitivity, centralization, and physical distance by phases. The pre-European group presents remarkable larger values for the transitivity, but lower values for the edges and centralization parameters compared to the post-European group. The distance parameters overlap between two groups. Both posterior estimates present generally symmetric distributions."}
knitr::include_graphics(here::here("analysis", "figures", "006-posterior-distribution.png"))
```

<!--posterior stats for for post-E and their comparison-->
After the arrival of Europeans, the nodal covariates of wealth rank show significant positive effects. This indicates burials in the same wealth level tend to be more connected with each other. Despite negative effects for ritual and sex variables, they are not significant. Also, there are no significant effects for variables of age and physical distance. Similar to the pre-European network, transitivity demonstrates a significant positive effect, but much weaker than in the pre-European network. In contrast, centralization has a significantly higher positive effect than the pre-European network. This means there is a tendency toward centralization that reflects a limited number of burials having many more ties than others. This implies more access to trade goods and the increased wealth accumulation and display behaviors in burial events. In general, the post-European network model has a smaller transitivity effect and a positive centralization effect compared to the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence.

@@

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand the robustness of comparison between two networks, we used a vertex bootstrap technique to cross-validate the results of our Bayesian ERGMs. The vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) to quantify standard errors and estimate sampling variability in the network statistics of interest [@Chen2019; @Snijders1999; @Rroberts2021]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks of different sizes by examining their confidence intervals for the network population. We used the vertex bootstrap to compute endogenous network statistics, including density, centralization, and transitivity for our two networks. We explored the sample size effect by removing nodes at certain percentages (5-40%) for both networks and comparing their confidence intervals. The results demonstrate a consistent difference between the two networks up to 30% node removals (Figure \@ref(fig:figure-bootstrap-CI)). This indicates that the network variables under investigation are robust to the different network size of our sample. Figure \@ref(fig:figure-bootstrap-CI) also shows consistency with our earlier finding of negative centralization in the pre-European period and positive centralization in the post-European period, over several conditions of node removal. Similarly consistent with our earlier result, the vertex bootstrap shows no significant difference on density and transitivity.

@@

<!--archaeological implication based on the results-->
Network covariates give insights into the socio-cultural factors that were associated with the acquisition and distribution of foreign prestige goods. Our results indicate that the major factor contributing to burial networks is ritual for the pre-European network, and wealth for the post-European network. We interpret these results as illustrating different social and economic mechanisms influencing relations between burials for the two periods. Before European arrival, ritual behaviors were important factors that structured the interconnections of prestige goods in the burial contexts. This might imply that ritual could be a status indicator at the time before foreign goods were first introduced to Indigenous societies. After the European presence was established, status indicators shifted from ritual to wealth differences. Moreover, geographical proximity, implying some kinship association, has not effect on sharing similar prestige items between burials for both phases. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based. Demographic information, such as age and sex, do not have any effects on network formation for both networks. But we also note that no effects could be due to the unavailability of age and sex values for many burials in our sample.
